pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-13 manuscript/greenwell.tex
minor edits

For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame. This is illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label. The result is displayed in the right side of Figure~\ref{fig:pd_lstat}. \strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code.
\begin{example}
boston.rf %>%
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
\strong{Note:} the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a third degree MARS model. \textit{Left}: Default plot. \textit{Middle}: Using a different color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}

@@

To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}. Note that while \code{xgboost} requires the training data to be a matrix (or an object of class \code{"dgCMatrix"} or \code{"xgb.DMatrix"}), \code{partial} requires the training data as a data frame with factors in tact!
\begin{example}
pred.list <- list("lstat", "rm", c("lstat", "rm"))
for (x in pred.list) {
  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
          chull = TRUE, train = subset(boston, select = -cmedv))
}
\end{example}

>>>>

For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame. This is illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label. The result is displayed in the right side of Figure~\ref{fig:pd_lstat}. \strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded for writing more readable code.
\begin{example}
boston.rf %>%  # the %>% operator is read as "and then"
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
\strong{Note:} the default color map for contour plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a third degree MARS model. \textit{Left}: Default plot. \textit{Middle}: Using a black and white color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}

@@

To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}. \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
pred.list <- list("lstat", "rm", c("lstat", "rm"))
for (x in pred.list) {
  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
          chull = TRUE, train = subset(boston, select = -cmedv))
}
\end{example}
